Hydra Battles in Coq

### 2020-11-10 doc/hydras.tex
Added multiplication of ordinal notations (first redaction)

\section{Examples of Ordinal Notations}

\subsection{The Ordinal \texorpdfstring{$\omega$}{omega}}

@@

\begin{Coqsrc}
Global Instance Omega : ON  Peano.lt Nat.compare.
Proof.
 split.
 - apply Nat.lt_strorder.
 - apply Wf_nat.lt_wf.
 - apply Nat.compare_spec.
Qed.
\end{Coqsrc}

\subsection{Sum of two Ordinal Notations}

@@

\emph{From Library~\href{https://coq.inria.fr/distrib/current/stdlib/Coq.Relations.Relation_Operators.html}{Relations.Relation\_Operators}}.

@@

@@

The ordinal $\omega^2$ (also called $\phi_0(2)$, see Chap.~\ref{chap:schutte}), is represented by the lexicographically ordered cartesian product $\mathbb{N}\times\mathbb{N}$.

\vspace{4pt}
\noindent\emph{From Module~\href{../theories/html/hydras.OrdinalNotations.ON_Omega2.html}{ON\_Omega2}}

\begin{Coqsrc}
Definition t := (nat * nat)%type.

Definition lt : relation t := lexico Peano.lt Peano.lt.
Definition le := clos_refl _ lt.

Infix "o<" := lt : o2_scope.
Infix "o<=" := le : o2_scope.
\end{Coqsrc}

It is easy to represent  finite ordinals, successors and limits.

@@

Notation "'F'" := fin : o2_scope.

@@

Definition succ (alpha : t) := (fst alpha, S (snd alpha)).

Notation "'omega'" := (1,0) : o2_scope.

@@

In order to build an ordinal notation out of the type \texttt{t}, we have to define a comparison function, and prove that the order \texttt{lt} is well-founded.

\begin{Coqsrc}
Instance lt_strorder : StrictOrder lt.
(* ... *)

Definition compare (alpha beta: t) : comparison :=
  match Nat.compare (fst alpha) (fst beta) with
    Eq => Nat.compare (snd alpha) (snd beta)
  | c => c
  end.

Instance Omega2: ON lt compare.
Proof.
  split.
  - apply lt_strorder.
  - apply lt_wf.
  - apply compare_correct.
Qed.
\end{Coqsrc}

@@

@@

We can define on \texttt{ON\_omega2} an addition which extends the addition on \texttt{nat}.

@@

 Context (Hvar : Hvariant lt_wf free m).
\end{Coqsrc}

@@

\index{Exercises}
\begin{exercise}

\label{sec:orgheadline63}
Write \emph{direct} proofs ({i.e.},  without applying the result and tools of Chap.~\ref{ks-chapter}) that the following data structures  are too simple for defining a variant for any hydra battle.

\begin{itemize}
\item  $\omega^n$ : the set of all $n$-uples of natural numbers, ordered  by 
  lexicographic ordering
\item  $\omega^\omega$: the set of all decreasing sequences (with respect to $\le$)  of natural numbers, ordered by lexicographic ordering on lists.

For instance, the following inequality holds:
\[\langle 4,3,3,3,3,3,3,2,2,2 \rangle\,<\,\langle 4,4,2 \rangle\]
\end{itemize}

@@

\index{Projects}
\label{exo:ON-mult}
\begin{project}
Define in \coq{} the product of two ordinal notations $N_A$ and $N_B$.
If $A$ [resp. $B$] is the underlying type of $N_A$ [resp. $N_B$], the
product \texttt{ON\_mult $N_A$ $N_B$} is implemented over the cartesian product $B\times A$ (with the lexicographic ordering).

@@

Prove that \texttt{ON\_mult (FinOrd $i$) Omega} is isomorphic to
\texttt{Omega}  whilst
\texttt{Omega}  is a sub-notation of \texttt{ON\_mult Omega (FinOrd $i$)},
for any strictly positive $i$.
\end{project}

>>>>

\section{Example: The Ordinal \texorpdfstring{$\omega$}{omega}}

@@

\begin{Coqsrc}
Global Instance Omega : ON  Peano.lt Nat.compare.
Proof.
 split.
 - apply Nat.lt_strorder.
 - apply Wf_nat.lt_wf.
 - apply Nat.compare_spec.
Qed.

Compute ON_compare 6 9.
\end{Coqsrc}

\section{Sum of  Ordinal Notations}

@@

\noindent
\emph{From Library~\href{https://coq.inria.fr/distrib/current/stdlib/Coq.Relations.Relation_Operators.html}{Relations.Relation\_Operators}}.

@@

\section{Product of Ordinal Notations}

Let \texttt{NA} and \texttt{NB} be two ordinal notations, on the respective  ordered types \texttt{A} and \texttt{B}. The product of \texttt{NA} and \texttt{NB} is considered as the concatenation of $B$ copies of $A$, ordered by the lexicographic order on $B\times A$.

\vspace{4pt}
\noindent
\emph{From Module~\href{../theories/html/hydras.OrdinalNotations.ON_mult.html}{OrdinalNotations.ON\_mult}}

Definition t := (B * A)%type.
Definition lt : relation t := lexico ltB ltA.
Definition le := clos_refl _ lt.

Definition compare (alpha beta: t) : comparison :=
  match compareB (fst alpha) (fst beta) with
  |  Eq => compareA (snd alpha) (snd beta)
  | c => c
  end.

Lemma compare_reflect alpha beta :
  match (compare alpha beta)
  with
    Lt => lt alpha  beta
  | Eq => alpha = beta
  | Gt => lt beta  alpha
  end.

Global Instance ON_mult : ON lt compare.

End Defs.
\end{Coqsrc}

@@

The ordinal $\omega^2$ (also called $\phi_0(2)$, see Chap.~\ref{chap:schutte}), is an instance of the multiplication presented in the last section.

\vspace{4pt}
\noindent\emph{From Module~\href{../theories/html/hydras.OrdinalNotations.ON_Omega2.html}{OrdinalNotations.ON\_Omega2}}

\begin{Coqsrc}
Definition Omega2 := ON_mult Omega Omega.
Existing Instance Omega2.
Definition t := ON_mult.t nat nat.
\end{Coqsrc}

@@

@@

Notation "'omega'" := (1,0) : ON_scope.

@@

@@

\subsubsection{Successor}

The successor of any ordinal is defined by a simple pattern-matching.
This function is proved to be correct w.r.t. the \texttt{Successor} predicate.

\begin{Coqsrc}
Definition succ (alpha : t) := (fst alpha, S (snd alpha)).
\end{Coqsrc}

\begin{Coqsrc}
Lemma lt_succ_le alpha beta : alpha o< beta <-> succ alpha o<= beta.

Lemma lt_succ alpha : alpha o< succ alpha.

Lemma succ_ok alpha beta : Successor beta alpha <-> beta = succ alpha.
\end{Coqsrc}

@@

We can define on \texttt{Omega2} an addition which extends the addition on \texttt{nat}.

@@

 Context (Hvar : Hvariant (Generic.wf (ON:=Omega2))  free m).
\end{Coqsrc}

@@

% \index{Exercises}
% \begin{exercise}

% \label{sec:orgheadline63}
% Write \emph{direct} proofs ({i.e.},  without applying the result and tools of Chap.~\ref{ks-chapter}) that the following data structures  are too simple for defining a variant for any hydra battle.

% \begin{itemize}
% \item  $\omega^n$ : the set of all $n$-uples of natural numbers, ordered  by 
%   lexicographic ordering
% \item  $\omega^\omega$: the set of all decreasing sequences (with respect to $\le$)  of natural numbers, ordered by lexicographic ordering on lists.

% For instance, the following inequality holds:
% \[\langle 4,3,3,3,3,3,3,2,2,2 \rangle\,<\,\langle 4,4,2 \rangle\]
% \end{itemize}

% \end{exercise}

@@

\index{Projects}
\label{exo:ON-mult}
\begin{project}
% Define in \coq{} the product of two ordinal notations $N_A$ and $N_B$.
% If $A$ [resp. $B$] is the underlying type of $N_A$ [resp. $N_B$], the
% product \texttt{ON\_mult $N_A$ $N_B$} is implemented over the cartesian product $B\times A$ (with the lexicographic ordering).

This exercise is about the non-commutativity of the multiplication of ordinals, reflected in ordinal notations.

@@

Prove formally  that \texttt{ON\_mult (FinOrd $i$) Omega} is isomorphic to
\texttt{Omega}  whilst
\texttt{Omega}  is a sub-notation of \texttt{ON\_mult Omega (FinOrd $i$)},
for any strictly positive $i$.
\end{project}
